\bibliographystyle{plain}
\begin{thebibliography}{56}

    \bibitem{sheldonaxler}
    \textsc{S. Axler},
    \textit{Linear Algebra Done Right},
    Springer,
    2015.

    \bibitem{higham}
    \textsc{N. J. Higham},
    \textit{Functions of Matrices: Theory and Computation},
    SIAM,
    2008.

    \bibitem{aurentzmachvandebrilwatkins}
    \textsc{J. L. Laurent, T. Mach, R. Vandebril, D. S. Watkins},
    \textit{Computing the eigenvalues of symmetric tridiagonal matrices via a Cayley transformation},
    Electronic Transactions on Numerical Analysis, 46: 447--459,
    2017.

    \bibitem{meerbergenspenceroose}
    \textsc{K. Meerbergen, A. Spence, D. Roose},
    \textit{Shift-invert and Cayley transforms for detection of rightmost eigenvalues of nonsymmetric matrices},
    Numerische Mathematik, 67(2): 161--189,
    1994.

    \bibitem{strichartz}
    \textsc{R. S. Strichartz},
    \textit{A Guide to Distribution Theory and Fourier Transforms},
    World Scientific,
    2003.

    \bibitem{richtmyer}
    \textsc{R. D. Richtmyer},
    \textit{Principles of Advanced Mathematical Physics},
    vol 1, Springer Verlag,
    New York,
    1981.

    \bibitem{reichl}
    \textsc{L. E. Reichl},
    \textit{A Modern Course in Statistical Physics},
    4th edition,
    Wiley-VCH,
    2016.

    \bibitem{sakurainapolitano}
    \textsc{J. J. Sakurai and J. Napolitano},
    \textit{Modern Quantum Mechanics},
    2nd edition,
    Addison-Wesley,
    2011.

    \bibitem{oppenheimschafer}
    \textsc{A. V. Oppenheim and R. W. Schafer},
    \textit{Discrete-Time Signal Processing},
    3rd edition,
    Prentice Hall,
    2010.

    \bibitem{mitra}
    \textsc{S. K. Mitra},
    \textit{Digital Signal Processing: A Computer-Based Approach},
    2nd edition,
    McGraw-Hill Science Engineering Math,
    2001.

    \bibitem{nortonkarczub}
    \textsc{M. P. Norton, D. G. Karczub},
    \textit{Fundamentals of Noise and Vibration Analysis for Engineers},
    Cambridge University Press,
    2003.

    \bibitem{szego1975}
    \textsc{G. Szeg\H{o}},
    \textit{Orthogonal Polynomials},
    4th edition,
    American Mathematical Society, Colloquium Publications, Vol. 23, 1975.

    \bibitem{masonhandscomb}
    \textsc{J. C. Mason and D. C. Handscomb},
    \textit{Chebyshev Polynomials},
    Chapman \& Hall/CRC,
    2003.

    \bibitem{weisse2006}
    \textsc{A. Weisse, G. Wellein, A. Alvermann, H. Fehske},
    \textit{The kernel polynomial method},
    Reviews of Modern Physics, 78(1): 275--306,
    2006.

    \bibitem{linsaadyang14}
    \textsc{L. Lin, Y. Saad, C. Yang},
    \textit{Approximating Spectral Densities of Large Matrices},
    SIAM Review, 58(1): 34--65,
    2016.

    \bibitem{golub2013matrix}
    \textsc{G. H. Golub and C. F. Van Loan},
    \textit{Matrix Computations},
    4th edition,
    Johns Hopkins University Press,
    Baltimore, MD,
    2013.

    \bibitem{sylvester1852}
    \textsc{J. J. Sylvester},
    \textit{A demonstration of the theorem that every homogeneous quadratic polynomial is reducible by real orthogonal substitutions to the form of a sum of positive and negative squares},
    Philosophical Magazine, 4(23): 138--142,
    1852.

    \bibitem{bennermach2012}
    \textsc{P. Benner and T. Mach},
    \textit{Computing all or some eigenvalues of symmetric \textnormal{$\mathcal{H}_l$}-matrices},
    SIAM Journal on Matrix Analysis and Applications, 33(4): 1395--1415,
    2012.

    \bibitem{hornjohnson2013}
    \textsc{R. A. Horn and C. R. Johnson},
    \textit{Matrix Analysis},
    2nd edition,
    Cambridge University Press,
    2013.

    \bibitem{kreyszig}
    \textsc{E. Kreyszig},
    \textit{Advanced Engineering Mathematics},
    10th edition,
    John Wiley \& Sons,
    Hoboken, NJ,
    2011.

    \bibitem{hutchinson1989}
    \textsc{M. F. Hutchinson},
    \textit{A Stochastic Estimator of the Trace of the Influence Matrix for Laplacian Smoothing Splines},
    Communications in Statistics - Simulation and Computation, 18(3): 1059--1076,
    1989.

    \bibitem{rudin1987}
    \textsc{W. Rudin},
    \textit{Real and Complex Analysis},
    3rd edition,
    McGraw-Hill,
    1987.

    \bibitem{trefethen1997}
    \textsc{L. N. Trefethen},
    \textit{Numerical Linear Algebra},
    SIAM,
    1997.

    \bibitem{saad2003}
    \textsc{Y. Saad},
    \textit{Iterative Methods for Sparse Linear Systems},
    2nd edition,
    SIAM,
    2003.

    \bibitem{mezzadri2007}
    \textsc{F. Mezzadri},
    \textit{How to generate random matrices from the classical compact groups},
    Notices of the AMS, 54(5), 592--604,
    2007.

    \bibitem{andersonguionnetzeitouni}
    \textsc{G. W. Anderson, A. Guionnet, O. Zeitouni},
    \textit{An Introduction to Random Matrices},
    Cambridge University Press,
    2010.

    \bibitem{PasSh11}
    \textsc{L. Pastur, M. Shcherbina},
    \textit{Eigenvalue Distribution of Large Random Matrices},
    AMS,
    2011.

    \bibitem{seidelessay2024}
    \textsc{C. Seidel},
    \textit{Annäherung an die Spektraldichte großer Matrizen},
    unpublished, Universität Potsdam, 2024.

\end{thebibliography}